---
title: "Data Visualization | Fall 2025 @AU"
subtitle: "Multivariate Analysis of GSS Data"
author: "Anne Skamris Holm and Nour Alsaid"
date: "`r Sys.Date()`"
header-includes:
- \usepackage{fancyhdr}
output:
  pdf_document:
    latex_engine: xelatex
    toc: true #Created table from headings
    number_sections: yes
  html_document:
    toc: yes
    number_sections: yes
    toc_float: yes
    theme: united
    highlight: espresso
  #always_allow_html: true
geometry: margin=1in
---

\pagestyle{fancy}
\fancyhead[HL]{Data Visualization} 
\fancyhead[HR]{Anne Skamris Holm and Nour Alsaid}

\newpage
```{r setup, include=FALSE}
# include = F results in coding chunks in doc not being printed
knitr::opts_chunk$set(echo = FALSE, include = TRUE, message = FALSE, warning = FALSE)

# Resetting the environment
rm(list=ls()) # clears global workspace

# Setting working directory
knitr::opts_knit$set(root.dir = '.')
getwd()

# Loading packages
pacman::p_load(tidyverse, pastecs, gridExtra, ggplot2, tinytex, shiny, bslib, plotly, webshot2, sf, jsonlite)

```

# Part 1

## Importing GSS data

```{r converting to CSV}
library(foreign)
  read.dct <- function(dct, labels.included = "yes") {
      temp <- readLines(dct)
      temp <- temp[grepl("_column", temp)]
      switch(labels.included,
             yes = {
                 pattern <- "_column\\(([0-9]+)\\)\\s+([a-z0-9]+)\\s+(.*)\\s+%([0-9]+)[a-z]\\s+(.*)"
                 classes <- c("numeric", "character", "character", "numeric", "character")
                 N <- 5
                 NAMES <- c("StartPos", "Str", "ColName", "ColWidth", "ColLabel")
             },
             no = {
                 pattern <- "_column\\(([0-9]+)\\)\\s+([a-z0-9]+)\\s+(.*)\\s+%([0-9]+).*"
                 classes <- c("numeric", "character", "character", "numeric")
                 N <- 4
                 NAMES <- c("StartPos", "Str", "ColName", "ColWidth")
             })
      temp_metadata <- setNames(lapply(1:N, function(x) {
          out <- gsub(pattern, paste("\\", x, sep = ""), temp)
          out <- gsub("^\\s+|\\s+$", "", out)
          out <- gsub('\"', "", out, fixed = TRUE)
          class(out) <- classes[x] ; out }), NAMES)
      temp_metadata[["ColName"]] <- make.names(gsub("\\s", "", temp_metadata[["ColName"]]))
      temp_metadata
  }

  read.dat <- function(dat, metadata_var, labels.included = "yes") {
      read.table(dat, col.names = metadata_var[["ColName"]])
  }


GSS_metadata <- read.dct("data/exploring_variables/GSS.dct")
GSS_ascii <- read.dat("data/exploring_variables/GSS.dat", GSS_metadata)
attr(GSS_ascii, "col.label") <- GSS_metadata[["ColLabel"]]
GSS <- GSS_ascii
  
# Save as csv
GSS %>% 
  write_csv(file = "data/exploring_variables/GSS_data.csv")

GSS_ascii %>% 
  write_csv(file = "data/exploring_variables/GSS_data_ascii.csv")

```

```{r}
# Showing headings and labels
sort(GSS_metadata[["ColName"]])
sort(GSS_metadata[["ColLabel"]])
```
## Preprocessing GSS data
```{r}
df_fresh <-read_csv('data/exploring_variables/GSS_data.csv')

#Insert row number ('ID') variable to the data frame
df_fresh <- dplyr::mutate(df_fresh, ID = row_number()) #generates student IDs by row number.

# Rename columns so they are all uppercase
df <- df_fresh %>%
  select(
    ID,
    GSS_ID=ID_,
    year=YEAR,
    sex_num=SEX,
    age=AGE,
    region=REGION,
    happiness=HAPPY,
    own_income=RINCOME,
    educ=EDUC,
    race=RACE,
    work=WRKSTAT,
    marital=MARITAL
)

df <- df %>%
  mutate(
    across(where(is.character), as.factor),
    ID = as.factor(ID),
    year = as.numeric(year),
    age = as.numeric(age),
    age= case_when(age < 0 ~ NA_real_, age>0~age),
    sex_num = as.factor(sex_num),
    sex = case_when(sex_num=="1"~"M", sex_num=="2"~"F", TRUE ~ NA_character_),
    region = as.factor(region),
    region = recode(region, "1" = "New England", "2" = "Middle Atlantic",  "3" = "East North Central",  "4" = "West North Central", "5" = "South Atlantic",  "6" = "East South Central",  "7" = "West South Central",  "8" = "Mountain",  "9" = "Pacific"),
    happiness = as.character(happiness),
    happiness = recode(happiness, "1" = "Very happy", "2" = "Pretty happy", "3" = "Not too happy", "-98" = "Do not know", "-100" = NA_character_, "-99" = NA_character_, "-97" = NA_character_),
    own_income = as.numeric(own_income),
    income_range_min = case_when(
      own_income < 1 ~ NA_real_,
      own_income %in% 1:4 ~ 0,
      own_income %in% 5:8 ~ 5000,
      own_income == 9    ~ 10000,
      own_income == 10   ~ 15000,
      own_income == 11   ~ 20000,
      own_income == 12   ~ 25000
    ),
    income_range_max = case_when(
      own_income < 1 ~ NA_real_,
      own_income %in% 1:4 ~ 4999,
      own_income %in% 5:8 ~ 9999,
      own_income == 9    ~ 14999,
      own_income == 10   ~ 19999,
      own_income == 11   ~ 24999,
      own_income == 12   ~ 99999
    ),

    income_range = case_when(
      own_income < 1 ~ NA_real_,
      own_income %in% 1:4 ~ 1,
      own_income %in% 5:8 ~ 2,
      own_income == 9    ~ 3,
      own_income == 10   ~ 4,
      own_income == 11   ~ 5,
      own_income == 12   ~ 6
    ),
    income_range = as.factor(income_range),

    race = as.character(race),
    race = recode(race, "-100" = NA_character_),
    educ = as.character(educ),
    educ = recode(educ, "-98" = NA_character_, "-99" = NA_character_),
    work = as.character(work),
    work = recode(work, "-97" = NA_character_, "-98" = NA_character_, "-99" = NA_character_),
    marital = as.character(marital),
    marital = recode(marital, "-97" = NA_character_, "-98" = NA_character_, "-99" = NA_character_)
  )

head(df)
```
```{r}
df_clean <- df %>%
  select(
    year,
    sex,
    age,
    region,
    happiness,
    income_range,
    income_range_min,
    income_range_max
)

data.frame(
  column = names(df_clean),
  na_count = colSums(is.na(df_clean)),
  na_percent = round(colSums(is.na(df_clean)) / nrow(df_clean) * 100, 1)
)

df_clean <- na.omit(df)
#df_clean <- df_clean %>%
#  filter(!is.na(age), !is.na(sex), !is.na(region))


data.frame(
  column = names(df_clean),
  na_count = colSums(is.na(df_clean)),
  na_percent = round(colSums(is.na(df_clean)) / nrow(df_clean) * 100, 1)
)

head(df_clean)
```

## Exploring GSS data
```{r}
ggplot(df_clean, aes(x = year)) + 
  theme_minimal() + 
  labs(x = "Income", title = "Income") + 
  geom_histogram()
```

# Part 2
## Building the shiny app
```{r}
# Creating df_clean with chosen variables
df_clean <- df %>%
  select(
    year,
    sex,
    age,
    region,
    happiness,
    #income_range,
    educ,
    race,
    work,
    marital
)

# Counting NA's in the full data
data.frame(
  #column = names(df_clean),
  na_count = colSums(is.na(df_clean)),
  na_percent = round(colSums(is.na(df_clean)) / nrow(df_clean) * 100, 1)
)

# Create simple missingness map
library(naniar)
vis_miss(df_clean)

# Creating df_NA consisting of the data removed
df_NA <- df_clean %>%
  filter(is.na(age) | is.na(sex) | is.na(region) | is.na(year) | is.na(happiness))

# Looking at filtered out data
data.frame(
  na_count = colSums(is.na(df_NA)),
  na_percent = round(colSums(is.na(df_NA)) / nrow(df_NA) * 100, 1)
)

# Removing NA's
df_clean <- na.omit(df_clean)
#df_clean <- df_clean %>%
#  filter(!is.na(age), !is.na(sex), !is.na(region))


# Counting NA's in df_clean (No NA's)
data.frame(
  #column = names(df_clean),
  na_count = colSums(is.na(df_clean)),
  na_percent = round(colSums(is.na(df_clean)) / nrow(df_clean) * 100, 1)
)

write.csv(df_clean, "data_vis_shiny_app/df_clean.csv", row.names = FALSE)
```

## Exploring GSS data

### Distribution of Data
```{r}
#YEAR
# Year including NA's
year_w_na <- ggplot(df, aes(x = year)) + 
  theme_minimal() + 
  labs(x = "Year", title = "Years with NA's") + 
  geom_bar(binwidth = 1)+
  coord_cartesian(ylim = c(0, 4000))  # sets y-axis range

# Year NA's removed
year_clean <- ggplot(df_clean, aes(x = year)) + 
  theme_minimal() + 
  labs(x = "Year", title = "Years with NA's removed") + 
  geom_bar(binwidth = 1)+
  coord_cartesian(ylim = c(0, 4000))  # sets y-axis range

# Year NA's 
year_na <- ggplot(df_NA, aes(x = year)) + 
  theme_minimal() + 
  labs(x = "Year", title = "Year removed responses") + 
  geom_bar(binwidth = 1)+
  coord_cartesian(ylim = c(0, 4000))  # sets y-axis range

#AGE
# age including NA's
age_w_na <- ggplot(df, aes(x = age)) + 
  theme_minimal() + 
  labs(x = "age", title = "ages with NA's") + 
  geom_bar(binwidth = 1)+
  coord_cartesian(ylim = c(0, 4000))  # sets y-axis range

# age NA's removed
age_clean <- ggplot(df_clean, aes(x = age)) + 
  theme_minimal() + 
  labs(x = "age", title = "ages with NA's removed") + 
  geom_bar(binwidth = 1)+
  coord_cartesian(ylim = c(0, 4000))  # sets y-axis range

# age NA's 
age_na <- ggplot(df_NA, aes(x = age)) + 
  theme_minimal() + 
  labs(x = "age", title = "age removed responses") + 
  geom_bar(binwidth = 1)+
  coord_cartesian(ylim = c(0, 4000))  # sets y-axis range

#REGION
# region including NA's
region_w_na <- ggplot(df, aes(x = region, fill=sex)) + 
  theme_minimal() + 
  labs(x = "region", title = "region with NA's") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 8000))+  # sets y-axis range
  coord_cartesian(xlim = c(0, 10))  # sets y-axis range

# region NA's removed
region_clean <- ggplot(df_clean, aes(x = region, fill=sex)) + 
  theme_minimal() + 
  labs(x = "region", title = "region groups with NA's removed") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 8000))  # sets y-axis range
 
# region NA's 
region_na <- ggplot(df_NA, aes(x = region, fill=sex)) + 
  theme_minimal() + 
  labs(x = "region", title = "region removed respones") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 8000))  # sets y-axis range

#SEX
# sex including NA's
sex_w_na <- ggplot(df, aes(x = sex, fill=sex)) + 
  theme_minimal() + 
  labs(x = "sex", title = "sex with NA's") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 40000))  # sets y-axis range

# sex NA's removed
sex_clean <- ggplot(df_clean, aes(x = sex, fill=sex)) + 
  theme_minimal() + 
  labs(x = "sex", title = "sex groups with NA's removed") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 40000))  # sets y-axis range
 
# sex NA's 
sex_na <- ggplot(df_NA, aes(x = sex, fill=sex)) + 
  theme_minimal() + 
  labs(x = "sex", title = "sex removed respones") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 40000))  # sets y-axis range


#HAPPINESS
# Happiness including NA's
happy_w_na <- ggplot(df, aes(x = happiness, fill=sex)) + 
  theme_minimal() + 
  labs(x = "Happiness", title = "Happiness with NA's") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 20000))  # sets y-axis range

# Happiness NA's removed
happy_clean <- ggplot(df_clean, aes(x = happiness, fill=sex)) + 
  theme_minimal() + 
  labs(x = "Happiness", title = "Happiness groups with NA's removed") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 20000))  # sets y-axis range

# Happiness NA's 
happy_na <- ggplot(df_NA, aes(x = happiness, fill=sex)) + 
  theme_minimal() + 
  labs(x = "Happiness", title = "Happiness removed respones") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 20000))  # sets y-axis range

# #INCOME
# # income including NA's
# income_w_na <- ggplot(df, aes(x = income_range, fill=sex)) + 
#   theme_minimal() + 
#   labs(x = "income", title = "income with NA's") + 
#   geom_bar(position = position_dodge(width = 0.9))+
#   coord_cartesian(ylim = c(0, 20000))  # sets y-axis range
# 
# # income NA's removed
# income_clean <- ggplot(df_clean, aes(x = income_range, fill=sex)) + 
#   theme_minimal() + 
#   labs(x = "income", title = "income groups with NA's removed") + 
#   geom_bar(position = position_dodge(width = 0.9))+
#   coord_cartesian(ylim = c(0, 20000))  # sets y-axis range
#  
# # income NA's 
# income_na <- ggplot(df_NA, aes(x = income_range, fill=sex)) + 
#   theme_minimal() + 
#   labs(x = "income", title = "income removed respones") + 
#   geom_bar(position = position_dodge(width = 0.9))+
#   coord_cartesian(ylim = c(0, 20000))  # sets y-axis range

#RACE
# race including NA's
race_w_na <- ggplot(df, aes(x = race, fill=sex)) + 
  theme_minimal() + 
  labs(x = "race", title = "race with NA's") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 35000))  # sets y-axis range

# race NA's removed
race_clean <- ggplot(df_clean, aes(x = race, fill=sex)) + 
  theme_minimal() + 
  labs(x = "race", title = "race groups with NA's removed") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 35000))  # sets y-axis range
 
# race NA's 
race_na <- ggplot(df_NA, aes(x = race, fill=sex)) + 
  theme_minimal() + 
  labs(x = "race", title = "race removed respones") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 35000))  # sets y-axis range


#MARITAL
# marital including NA's
marital_w_na <- ggplot(df, aes(x = marital, fill=sex)) + 
  theme_minimal() + 
  labs(x = "marital", title = "marital with NA's") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 21000))  # sets y-axis range

# marital NA's removed
marital_clean <- ggplot(df_clean, aes(x = marital, fill=sex)) + 
  theme_minimal() + 
  labs(x = "marital", title = "marital groups with NA's removed") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 21000))  # sets y-axis range
 
# marital NA's 
marital_na <- ggplot(df_NA, aes(x = marital, fill=sex)) + 
  theme_minimal() + 
  labs(x = "marital", title = "marital removed respones") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 21000))  # sets y-axis range

#EDUCATION
# educ including NA's
educ_w_na <- ggplot(df, aes(x = educ, fill=sex)) + 
  theme_minimal() + 
  labs(x = "educ", title = "educ with NA's") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 15000))  # sets y-axis range

# educ NA's removed
educ_clean <- ggplot(df_clean, aes(x = educ, fill=sex)) + 
  theme_minimal() + 
  labs(x = "educ", title = "educ groups with NA's removed") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 15000))  # sets y-axis range
 
# educ NA's 
educ_na <- ggplot(df_NA, aes(x = educ, fill=sex)) + 
  theme_minimal() + 
  labs(x = "educ", title = "educ removed respones") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 15000))  # sets y-axis range

#WORK
# work including NA's
work_w_na <- ggplot(df, aes(x = work, fill=sex)) + 
  theme_minimal() + 
  labs(x = "work", title = "work with NA's") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 20000))  # sets y-axis range

# work NA's removed
work_clean <- ggplot(df_clean, aes(x = work, fill=sex)) + 
  theme_minimal() + 
  labs(x = "work", title = "work groups with NA's removed") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 20000))  # sets y-axis range
 
# work NA's 
work_na <- ggplot(df_NA, aes(x = work, fill=sex)) + 
  theme_minimal() + 
  labs(x = "work", title = "work removed respones") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 20000))  # sets y-axis range

grid.arrange(year_w_na, year_clean, year_na, ncol=2)
grid.arrange(age_w_na, age_clean, age_na, ncol=2)
grid.arrange(region_w_na, region_clean, region_na, ncol=2)
grid.arrange(sex_w_na, sex_clean, sex_na, ncol=2)
grid.arrange(happy_w_na, happy_clean, happy_na, ncol=2)
#grid.arrange(income_w_na, income_clean, income_na, ncol=2)
grid.arrange(race_w_na, race_clean, race_na, ncol=2)
grid.arrange(marital_w_na, marital_clean, marital_na, ncol=2)
grid.arrange(educ_w_na, educ_clean, educ_na, ncol=2)
grid.arrange(work_w_na, work_clean, work_na, ncol=2)
```


### Variable distributions percentage
```{r}
# happiness 
df_clean %>%
  group_by(sex, happiness) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(sex) %>%                        # calculate percentages within each sex
  mutate(perc = count / sum(count) * 100) %>%
  
  ggplot(aes(x = happiness, y = perc, fill = sex)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  theme_minimal() +
  labs(x = "Happiness", y = "Percentage", title = "Happiness distribution") +
  scale_y_continuous(limits = c(0, 100))

```


### Variables across time
```{r}
# Happiness across Time
library(dplyr)
library(ggplot2)

  ggplot(df_clean %>%
    group_by(year, happiness) %>%
    summarise(count = n(), .groups = "drop") %>%
    group_by(year) %>%
    mutate(perc = count / sum(count) * 100),
  aes(x = year, y = perc, color = happiness, group = happiness)
) +
  geom_point(size = 2) +
  geom_line() +
  theme_minimal() +
  labs(
    x = "Year",
    y = "Percentage of Responses",
    color = "Happiness",
    title = "Happiness Distribution across Time"
  )

```

```{r}
# # Income across Time
# library(dplyr)
# library(ggplot2)
# 
# # count number of responses per year × income
# df_counts <- df_clean %>%
#   group_by(year, income_range) %>%
#   summarise(count = n(), .groups = "drop")
# 
# # connected scatterplot
# ggplot(df_counts, aes(x = year, y = count, color = income_range, group = income_range)) +
#   geom_point(size = 2) +
#   geom_line() +
#   theme_minimal() +
#   labs(
#     x = "Year",
#     y = "Number of Responses",
#     color = "Income",
#     title = "Income Counts across Time"
#   )
# 
# # percentage of each income group per year
# df_perc <- df_clean %>%
#   group_by(year, income_range) %>%
#   summarise(count = n(), .groups = "drop") %>%
#   group_by(year) %>%
#   mutate(perc = count / sum(count) * 100)
# 
# # connected scatterplot
# ggplot(df_perc, aes(x = year, y = perc, color = income_range, group = income_range)) +
#   geom_point(size = 2) +
#   geom_line() +
#   theme_minimal() +
#   labs(
#     x = "Year",
#     y = "Percentage of Responses",
#     color = "Income",
#     title = "Income Distribution across Time"
#   )
```


### Variables across region
`### Variables across region
```{r}
# Happiness
ggplot(df_clean, aes(x = interaction(happiness, region), fill = sex)) + 
  geom_bar(position = "stack") +
  theme_minimal() +
  labs(
    x = "Happiness and Region",
    y = "Count",
    fill = "Sex",
    title = "Counts of responses for Happiness by Regions and Sex"
  )+
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1) # rotate labels
  )

# # Income
# ggplot(df_clean, aes(x = interaction(income_range, region), fill = sex)) + 
#   geom_bar(position = "stack") +
#   theme_minimal() +
#   labs(
#     x = "Income and Region",
#     y = "Count",
#     fill = "Sex",
#     title = "Counts of responses for Income by Regions and Sex"
#   )+
#   theme(
#     axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1) # rotate labels
#   )



```
```{r}
# Define mapping of regions to state abbreviations
us_regions <- list(
  "New England" = c("CT", "ME", "MA", "NH", "RI", "VT"),
  "Middle Atlantic" = c("NJ", "NY", "PA"),
  "East North Central" = c("IL", "IN", "MI", "OH", "WI"),
  "West North Central" = c("IA", "KS", "MN", "MO", "NE", "ND", "SD"),
  "South Atlantic" = c("DE", "DC", "FL", "GA", "MD", "NC", "SC", "VA", "WV"),
  "East South Central" = c("AL", "KY", "MS", "TN"),
  "West South Central" = c("AR", "LA", "OK", "TX"),
  "Mountain" = c("AZ", "CO", "ID", "MT", "NV", "NM", "UT", "WY"),
  "Pacific" = c("AK", "CA", "HI", "OR", "WA")
)

# Convert the list to a tidy dataframe
region_to_state_df <- tibble(
  region = rep(names(us_regions), lengths(us_regions)),
  state = unlist(us_regions)
)

# Assign the data 
data_df <- data.frame(
  region = c(
    "New England",
    "Middle Atlantic",
    "East North Central",
    "West North Central",
    "South Atlantic",
    "East South Central",
    "West South Central",
    "Mountain",
    "Pacific"
  ),
  data = c(2, 6, 15, 13, 20, 3, 5, 7, 12)
)


```

```{r}
library(sf)
library(leaflet)

# Loading map
my_map <- read_sf("data_vis_shiny_app/us_states/cb_2024_us_state_500k.shp")
my_map <- my_map %>%
  rename(state=STUSPS)%>%
  filter(!state %in% c("AS", "GU", "MP", "PR", "VI"))%>%
  left_join(region_to_state_df %>% select(state, region), by = "state")

#Converting states to regions
my_map <- my_map %>%
  group_by(region) %>%
  summarise(geometry = st_union(geometry))%>%  # merge polygons
  left_join(data_df %>% select(data, region), by = "region")


# Create a color palette for the map:
region_palette <- colorFactor(
  palette = "viridis", 
  domain = my_map$region, 
  na.color = "transparent")
```


```{r}
ggplot(df_clean, aes(x = educ, fill=sex)) + 
    theme_minimal() + 
    labs(x = "educ", title = "Education Groups") + 
    geom_bar(position = position_dodge(width = 0.9))+
    coord_cartesian(ylim = c(0, 15000))
```

# Part 2
## Functions to use in app.R

```{r}
sum(is.na(df_clean$sex))
sum(is.na(df_clean$educ))
```


```{r}
# Education Groups:
p_educ <-ggplot(df_clean %>%
             group_by(sex, educ) %>% 
             summarise(count = n(), .groups = "drop") %>% 
             group_by(educ) %>%
             mutate(perc = count / sum(count) * 100),
           aes(x = educ, y=perc, fill=sex)) + 
      geom_col(position = position_dodge(width = 0.9))+
      theme_minimal() + 
      #theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))+
      labs(title = "Education Groups", fill="Gender") + 
      coord_cartesian(ylim = c(0, 100))


# Happiness Groups
p_hap <- ggplot(df_clean%>%
         group_by(sex, happiness) %>% 
          summarise(count = n(), .groups = "drop") %>% 
          group_by(happiness) %>%
          mutate(perc = count / sum(count) * 100),
          aes(x = happiness, y=perc, fill=sex)) + 
    geom_col(position = position_dodge(width = 0.9))+
    #theme_minimal() + 
      #https://ggplot2.tidyverse.org/reference/element.html
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=.95, size = rel(.80)))+
    labs(title = "Happiness Groups", fill="Gender") + 
    coord_cartesian(ylim = c(0, 100))
    


# Race Groups
p_race <- ggplot(df_clean%>%
             group_by(sex, race) %>% 
             summarise(count = n(), .groups = "drop") %>% 
             group_by(race) %>%
             mutate(perc = count / sum(count) * 100),
           aes(x = race, y=perc, fill=sex)) + 
    geom_col(position = position_dodge(width = 0.9))+
      theme_minimal() + 
      labs(x = "race", title = "Race Groups") + 
      coord_cartesian(ylim = c(0, 100))

p_marital <- ggplot(df_clean%>%
             group_by(sex, marital) %>% 
             summarise(count = n(), .groups = "drop") %>% 
             group_by(marital) %>%
             mutate(perc = count / sum(count) * 100),
           aes(x = marital, y=perc, fill=sex)) + 
    geom_col(position = position_dodge(width = 0.9))+
      theme_minimal() + 
      labs(x = "marital", title = "Marraige Type") + 
      coord_cartesian(ylim = c(0, 100))
  
p_work <- ggplot(df_clean%>%
             group_by(sex, work) %>% 
             summarise(count = n(), .groups = "drop") %>% 
             group_by(work) %>%
             mutate(perc = count / sum(count) * 100),
           aes(x = work, y=perc, fill=sex)) + 
    geom_col(position = position_dodge(width = 0.9))+
      theme_minimal() + 
      labs(x = "work", title = "Working Classes") + 
      coord_cartesian(ylim = c(0, 100))



all_ages <- unique(df_clean$age)
#p_age <- ggplot(df_clean%>%
#             group_by(sex, age) %>% 
#             summarise(count = n(), .groups = "drop") %>% 
#             group_by(age) %>%
#             mutate(perc = count / sum(count) * 100),
#           aes(x = age, y=perc, fill=sex)) + 
#    geom_area(alpha=0.8, size=.5, colour="white")
#      #theme_minimal() + 
#      theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust=.95, size = rel(.80)))+
#      labs(x = "age", title = "Sex responses' Age")+
#     coord_cartesian(ylim = c(0, 100))

diverging_age <- df_clean %>%
  group_by(sex, age) %>% 
  summarise(count = n(), .groups = "drop") %>% 
  group_by(age) %>%
  mutate(perc = count / sum(count) * 100) %>%
  mutate(perc_diverging = ifelse(sex == "F", -perc, perc))
p_age <- ggplot(diverging_age,
       aes(x = age, y = perc_diverging, fill = sex)) + 
  geom_col() +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1, size = rel(.80))) +
  labs(x = "Age", y = "Percentage", title = "Sex responses per Age") +
  coord_cartesian(ylim = c(-100, 100))+
  scale_x_continuous(breaks = all_ages) +
  scale_y_continuous(labels = function(x) abs(x))
  #coord_flip()
  


#p_year <- ggplot(df_clean%>%
#             group_by(sex, year) %>% 
#             summarise(count = n(), .groups = "drop") %>% 
#             group_by(year) %>%
#             mutate(perc = count / sum(count) * 100),
#           aes(x = year, y=perc, fill=sex)) + 
#    geom_col(position = position_dodge(width = 0.9))+
      #theme_minimal() + 
#      theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust=.95, size = rel(.80)))+
#      labs(x = "Year", title = "Sex responses per Year") +
#      coord_cartesian(ylim = c(0, 100))+
#      scale_x_continuous(breaks=complete_years)

complete_years <- min(df_clean$year):max(df_clean$year)
diverging_year <- df_clean %>%
  group_by(sex, year) %>% 
  summarise(count = n(), .groups = "drop") %>% 
  group_by(year) %>%
  mutate(perc = count / sum(count) * 100) %>%
  mutate(perc_diverging = ifelse(sex == "F", -perc, perc))

# Then plot
p_year <- ggplot(diverging_year,
       aes(x = year, y = perc_diverging, fill = sex)) + 
  geom_col() +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1, size = rel(.80))) +
  labs(x = "Year", y = "Percentage", title = "Sex responses per Year") +
  coord_cartesian(ylim = c(-100, 100))+
  scale_x_continuous(breaks = complete_years) +
  scale_y_continuous(labels = function(x) abs(x))+
  coord_flip()
  


p_educ
p_hap
p_race
p_marital
p_work
p_age
p_year
```





# Part 2
## Building the shiny app
```{r}
library(shiny)
# runApp("data_vis_shiny_app") #Shows only the app and no code

#runApp("data_vis_shiny_app", display.mode = "showcase") #Shows the code as well and highlights chinks with interaction
```