---
title: "Data Visualization | Fall 2025 @AU"
subtitle: "Multivariate Analysis of GSS Data"
author: "Anne Skamris Holm and Nour Alsaid"
date: "`r Sys.Date()`"
header-includes:
- \usepackage{fancyhdr}
output:
  html_document:
    toc: yes
    number_sections: yes
    toc_float: yes
    theme: united
    highlight: espresso
  pdf_document:
    latex_engine: xelatex
    toc: true #Created table from headings
    number_sections: yes
  #always_allow_html: true
geometry: margin=1in
---

\pagestyle{fancy}
\fancyhead[HL]{Data Visualization} 
\fancyhead[HR]{Anne Skamris Holm and Nour Alsaid}

\newpage
```{r setup, include=FALSE}
# include = F results in coding chunks in doc not being printed
knitr::opts_chunk$set(echo = FALSE, include = TRUE, message = FALSE, warning = FALSE)

# Resetting the environment
rm(list=ls()) # clears global workspace

# Setting working directory
knitr::opts_knit$set(root.dir = '.')
getwd()

# Loading packages
pacman::p_load(tidyverse, pastecs, gridExtra, ggplot2, tinytex, shiny, bslib, plotly, webshot2, sf, jsonlite)

```

# Part 1

## Importing GSS data

```{r converting to CSV}
library(foreign)
  read.dct <- function(dct, labels.included = "yes") {
      temp <- readLines(dct)
      temp <- temp[grepl("_column", temp)]
      switch(labels.included,
             yes = {
                 pattern <- "_column\\(([0-9]+)\\)\\s+([a-z0-9]+)\\s+(.*)\\s+%([0-9]+)[a-z]\\s+(.*)"
                 classes <- c("numeric", "character", "character", "numeric", "character")
                 N <- 5
                 NAMES <- c("StartPos", "Str", "ColName", "ColWidth", "ColLabel")
             },
             no = {
                 pattern <- "_column\\(([0-9]+)\\)\\s+([a-z0-9]+)\\s+(.*)\\s+%([0-9]+).*"
                 classes <- c("numeric", "character", "character", "numeric")
                 N <- 4
                 NAMES <- c("StartPos", "Str", "ColName", "ColWidth")
             })
      temp_metadata <- setNames(lapply(1:N, function(x) {
          out <- gsub(pattern, paste("\\", x, sep = ""), temp)
          out <- gsub("^\\s+|\\s+$", "", out)
          out <- gsub('\"', "", out, fixed = TRUE)
          class(out) <- classes[x] ; out }), NAMES)
      temp_metadata[["ColName"]] <- make.names(gsub("\\s", "", temp_metadata[["ColName"]]))
      temp_metadata
  }

  read.dat <- function(dat, metadata_var, labels.included = "yes") {
      read.table(dat, col.names = metadata_var[["ColName"]])
  }


GSS_metadata <- read.dct("data/exploring_variables/GSS.dct")
GSS_ascii <- read.dat("data/exploring_variables/GSS.dat", GSS_metadata)
attr(GSS_ascii, "col.label") <- GSS_metadata[["ColLabel"]]
GSS <- GSS_ascii
  
# Save as csv
GSS %>% 
  write_csv(file = "data/exploring_variables/GSS_data.csv")

GSS_ascii %>% 
  write_csv(file = "data/exploring_variables/GSS_data_ascii.csv")

```




## Preprocessing GSS data
```{r}
df_fresh <-read_csv('data/exploring_variables/GSS_data.csv')

#Insert row number ('ID') variable to the data frame
df_fresh <- dplyr::mutate(df_fresh, ID = row_number()) #generates student IDs by row number.

# Rename columns so they are all uppercase
df <- df_fresh %>%
  select(
    ID,
    GSS_ID=ID_,
    year=YEAR,
    sex_num=SEX,
    age=AGE,
    region=REGION,
    happiness=HAPPY,
    educ=EDUC,
    race=RACE,
    work=WRKSTAT,
    marital=MARITAL
)

df <- df %>%
  mutate(
    across(where(is.character), as.factor),
    ID = as.factor(ID),
    
    year = as.numeric(year),
    
    age = as.numeric(age),
    age= case_when(age < 0 ~ NA_real_, age>0~age),
    
    sex_num = as.factor(sex_num),
    sex = case_when(sex_num=="1"~"M", sex_num=="2"~"F", TRUE ~ NA_character_),
    
    region = as.factor(region),
    region = recode(region, "1" = "New England", "2" = "Middle Atlantic",  "3" = "East North Central",  "4" = "West North Central", "5" = "South Atlantic",  "6" = "East South Central",  "7" = "West South Central",  "8" = "Mountain",  "9" = "Pacific"),
    
    happiness = as.character(happiness),
    happiness = recode(happiness, "1" = "Very happy", "2" = "Pretty happy", "3" = "Not too happy", "-98" = NA_character_, "-100" = NA_character_, "-99" = NA_character_, "-97" = NA_character_),
    
    race = as.character(race),
    race = recode(race, "-100" = NA_character_),
    race = recode(race, "1" = "White", "2" = "Black", "3" = "Other"),
    
    
    educ = ifelse(educ %in% c(-99, -98), NA, educ),
    educ = cut(educ, 
    breaks = c(0, 1, 9, 13, 15, 17, 19, Inf),
    labels = c("Kindergarten", "Elementary School", 
               "High School", "College Degree", 
               "Bachelor's Degree", "Master's Degree", "Advanced Professional Degree"),
    right = FALSE),
    
    
    work = as.character(work),
    work = recode(work, "-97" = NA_character_, "-98" = NA_character_, "-99" = NA_character_),
    work = recode(work, "1" = "Working full time", "2" = "Working part time", "3" = "With a job, but home", "4" = "Unemployed", "5" = "Retired", "6" = "In school", "7" = "Keeping house", "8" = "Other"),
    
    marital = as.character(marital),
    marital = recode(marital, "-97" = NA_character_, "-98" = NA_character_, "-99" = NA_character_),
    marital = recode(marital, "1" = "Married", "2" = "Widowed", "3" = "Divorced", "4" = "Sepreated", "5" = "Never married")
  )

head(df)
```

```{r}
unique(df$educ)
```


### cleaning the data, eg. removing NA's
```{r}
unique(df$region)

# Creating df_clean with chosen variables
df_clean <- df %>%
  select(
    year,
    sex,
    age,
    region,
    happiness,
    educ,
    race,
    work,
    marital
)

# Counting NA's in the full data
data.frame(
  na_count = colSums(is.na(df_clean)),
  na_percent = round(colSums(is.na(df_clean)) / nrow(df_clean) * 100, 1)
)

# Create simple missingness map
library(naniar)
vis_miss(df_clean)

# Creating df_NA consisting of the data removed
df_NA <- df_clean %>%
  filter(is.na(age) | is.na(sex) | is.na(region) | is.na(year) | is.na(happiness))


# Removing NA's
df_clean <- na.omit(df_clean)
#df_clean <- df_clean %>%
#  filter(!is.na(age), !is.na(sex), !is.na(region))


# Adding age ranges after cleaning

# Source - https://stackoverflow.com/questions/51095156/how-to-group-ages-in-specific-range-using-r
# Posted by Rui Barradas
# Retrieved 2025-11-12, License - CC BY-SA 4.0
labelsL = c("18 to 19", "20 to 24", "25 to 29", "30 to 34", "35 to 39", "40 to 44", "45 to 49", "50 to 54", "55 to 59", "60 to 64", "65 to 69", "70 to 74", "75 to 79", "80 to 84", "85 to 89")
age_ranges <- cut(df_clean$age, c(17, seq(19, 89, by = 5)), labels = labelsL, Right=FALSE)

df_clean$age_ranges <- age_ranges


list_year <- labelsL <- c("1970's","1980's","1990's","2000's","2010's","2020's")

year_ranges <-cut(df_clean$year, c(1970, 1979 ,1989, 1999, 2009, 2019, 2029) , labels = list_year, include.lowest = TRUE)

df_clean$year_ranges <- year_ranges



# Counting NA's in df_clean (No NA's)
data.frame(
  #column = names(df_clean),
  na_count = colSums(is.na(df_clean)),
  na_percent = round(colSums(is.na(df_clean)) / nrow(df_clean) * 100, 1)
)

write.csv(df_clean, "data_vis_shiny_app/df_clean.csv", row.names = FALSE)
```

```{r}
l = c(seq(18, 89, by = 1))
labelsL = c("18 to 19", "20 to 24", "25 to 29", "30 to 34", "35 to 39", "40 to 44", "45 to 49", "50 to 54", "55 to 59", "60 to 64", "65 to 69", "70 to 74", "75 to 79", "80 to 84", "85 to 89")
cut(l, c(17, seq(19, 89, by = 5)), labels = labelsL, Right=FALSE)
l[8]
```
```{r}
labelsL <- c("1970's","1980's","1990's","2000's","2010's","2020's")

decades <- c(1970:1979, 1980:1989, 1990:1999, 2000:2009, 2010:2019, 2020:2029)
c(1970, 1980 ,1990, 2000, 2010, 2020, 2029)

cut(decades, c(1970, 1979 ,1989, 1999, 2009, 2019, 2029) , labels = labelsL, include.lowest = TRUE)
decades[11]
```


## Exploring GSS data

### Distribution of Data
```{r}
#YEAR
# Year including NA's
year_w_na <- ggplot(df, aes(x = year)) + 
  theme_minimal() + 
  labs(x = "Year", title = "Years with NA's") + 
  geom_bar(binwidth = 1)+
  coord_cartesian(ylim = c(0, 4000))  # sets y-axis range

# Year NA's removed
year_clean <- ggplot(df_clean, aes(x = year)) + 
  theme_minimal() + 
  labs(x = "Year", title = "Years with NA's removed") + 
  geom_bar(binwidth = 1)+
  coord_cartesian(ylim = c(0, 4000))  # sets y-axis range

# Year NA's 
year_na <- ggplot(df_NA, aes(x = year)) + 
  theme_minimal() + 
  labs(x = "Year", title = "Year removed responses") + 
  geom_bar(binwidth = 1)+
  coord_cartesian(ylim = c(0, 4000))  # sets y-axis range

#AGE
# age including NA's
age_w_na <- ggplot(df, aes(x = age)) + 
  theme_minimal() + 
  labs(x = "age", title = "ages with NA's") + 
  geom_bar(binwidth = 1)+
  coord_cartesian(ylim = c(0, 4000))  # sets y-axis range

# age NA's removed
age_clean <- ggplot(df_clean, aes(x = age)) + 
  theme_minimal() + 
  labs(x = "age", title = "ages with NA's removed") + 
  geom_bar(binwidth = 1)+
  coord_cartesian(ylim = c(0, 4000))  # sets y-axis range

# age NA's 
age_na <- ggplot(df_NA, aes(x = age)) + 
  theme_minimal() + 
  labs(x = "age", title = "age removed responses") + 
  geom_bar(binwidth = 1)+
  coord_cartesian(ylim = c(0, 4000))  # sets y-axis range

#REGION
# region including NA's
region_w_na <- ggplot(df, aes(x = region, fill=sex)) + 
  theme_minimal() + 
  labs(x = "region", title = "region with NA's") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 15000))  # sets y-axis range

# region NA's removed
region_clean <- ggplot(df_clean, aes(x = region, fill=sex)) + 
  theme_minimal() + 
  labs(x = "region", title = "region groups with NA's removed") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 15000))  # sets y-axis range
 
# region NA's 
region_na <- ggplot(df_NA, aes(x = region, fill=sex)) + 
  theme_minimal() + 
  labs(x = "region", title = "region removed respones") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 15000))  # sets y-axis range

#SEX
# sex including NA's
sex_w_na <- ggplot(df, aes(x = sex, fill=sex)) + 
  theme_minimal() + 
  labs(x = "sex", title = "sex with NA's") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 40000))  # sets y-axis range

# sex NA's removed
sex_clean <- ggplot(df_clean, aes(x = sex, fill=sex)) + 
  theme_minimal() + 
  labs(x = "sex", title = "sex groups with NA's removed") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 40000))  # sets y-axis range
 
# sex NA's 
sex_na <- ggplot(df_NA, aes(x = sex, fill=sex)) + 
  theme_minimal() + 
  labs(x = "sex", title = "sex removed respones") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 40000))  # sets y-axis range


#HAPPINESS
# Happiness including NA's
happy_w_na <- ggplot(df, aes(x = happiness, fill=sex)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1, size = rel(.80))) + 
  labs(x = "Happiness", title = "Happiness with NA's") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 20000))  # sets y-axis range

# Happiness NA's removed
happy_clean <- ggplot(df_clean, aes(x = happiness, fill=sex)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1, size = rel(.80))) + 
  labs(x = "Happiness", title = "Happiness groups with NA's removed") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 20000))  # sets y-axis range

# Happiness NA's 
happy_na <- ggplot(df_NA, aes(x = happiness, fill=sex)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1, size = rel(.80))) + 
  labs(x = "Happiness", title = "Happiness removed respones") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 20000))  # sets y-axis range

#RACE
# race including NA's
race_w_na <- ggplot(df, aes(x = race, fill=sex)) + 
  theme_minimal() + 
  labs(x = "race", title = "race with NA's") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 35000))  # sets y-axis range

# race NA's removed
race_clean <- ggplot(df_clean, aes(x = race, fill=sex)) + 
  theme_minimal() + 
  labs(x = "race", title = "race groups with NA's removed") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 35000))  # sets y-axis range
 
# race NA's 
race_na <- ggplot(df_NA, aes(x = race, fill=sex)) + 
  theme_minimal() + 
  labs(x = "race", title = "race removed respones") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 35000))  # sets y-axis range

#MARITAL
# marital including NA's
marital_w_na <- ggplot(df, aes(x = marital, fill=sex)) + 
  theme_minimal() + 
  labs(x = "marital", title = "marital with NA's") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 21000))  # sets y-axis range

# marital NA's removed
marital_clean <- ggplot(df_clean, aes(x = marital, fill=sex)) + 
  theme_minimal() + 
  labs(x = "marital", title = "marital groups with NA's removed") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 21000))  # sets y-axis range
 
# marital NA's 
marital_na <- ggplot(df_NA, aes(x = marital, fill=sex)) + 
  theme_minimal() + 
  labs(x = "marital", title = "marital removed respones") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 21000))  # sets y-axis range

#EDUCATION
# educ including NA's
educ_w_na <- ggplot(df, aes(x = educ, fill=sex)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1, size = rel(.80))) + 
  labs(x = "educ", title = "educ with NA's") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 15000))  # sets y-axis range

# educ NA's removed
educ_clean <- ggplot(df_clean, aes(x = educ, fill=sex)) + 
  labs(x = "educ", title = "educ groups with NA's removed") + 
  geom_bar(position = position_dodge(width = 0.9))+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1, size = rel(.80)))+
  coord_cartesian(ylim = c(0, 15000))  # sets y-axis range
 
# educ NA's 
educ_na <- ggplot(df_NA, aes(x = educ, fill=sex)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1, size = rel(.80))) + 
  labs(x = "educ", title = "educ removed respones") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 15000))  # sets y-axis range

#WORK
# work including NA's
work_w_na <- ggplot(df, aes(x = work, fill=sex)) + 
  theme_minimal() + 
  labs(x = "work", title = "work with NA's") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 20000))  # sets y-axis range

# work NA's removed
work_clean <- ggplot(df_clean, aes(x = work, fill=sex)) + 
  theme_minimal() + 
  labs(x = "work", title = "work groups with NA's removed") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 20000))  # sets y-axis range
 
# work NA's 
work_na <- ggplot(df_NA, aes(x = work, fill=sex)) + 
  theme_minimal()+ 
  labs(x = "work", title = "work removed respones") + 
  geom_bar(position = position_dodge(width = 0.9))+
  coord_cartesian(ylim = c(0, 20000))  # sets y-axis range

# grid.arrange(year_w_na, year_clean, year_na, ncol=2)
# grid.arrange(age_w_na, age_clean, age_na, ncol=2)
# grid.arrange(region_w_na, region_clean, region_na, ncol=2)
# grid.arrange(sex_w_na, sex_clean, sex_na, ncol=2)
# grid.arrange(happy_w_na, happy_clean, happy_na, ncol=2)
# grid.arrange(race_w_na, race_clean, race_na, ncol=2)
# grid.arrange(marital_w_na, marital_clean, marital_na, ncol=2)
# grid.arrange(educ_w_na, educ_clean, educ_na, ncol=2)
# grid.arrange(work_w_na, work_clean, work_na, ncol=2)
```

# Part 2 -building the app

### Creating data files
```{r}
# Assign the data 
library(dplyr)
library(readr)

# Define your decades and regions
decades <- list(
  "1970's" = 1970:1979,
  "1980's" = 1980:1989,
  "1990's" = 1990:1999,
  "2000's" = 2000:2009,
  "2010's" = 2010:2019,
  "2020's" = 2020:2029
)

regions <- c(
  "New England", "Middle Atlantic", "East North Central", "West North Central"
  #, "South Atlantic", "East South Central", "West South Central", "Mountain", "Pacific"
)

# Create a complete grid of all combinations (even if missing in data)
df_happy <- expand_grid(
  sex = unique(df_clean$sex),
  happiness = unique(df_clean$happiness),
  region = regions,
  year = names(decades)
)

write.csv(df_happy, "data_vis_shiny_app/data_happy.csv", row.names = FALSE)

# Create a complete grid of all combinations of educ
df_educ <- expand_grid(
  sex = unique(df_clean$sex),
  educ = unique(df_clean$educ),
  region = regions,
  year = names(decades)
)

write.csv(df_educ, "data_vis_shiny_app/data_educ.csv", row.names = FALSE)
unique(df_educ$educ)

# Create a complete grid of all combinations of educ
df_marital <- expand_grid(
  sex = unique(df_clean$sex),
  marital = unique(df_clean$marital),
  region = regions,
  year = names(decades)
)
unique(df_marital$marital)

write.csv(df_marital, "data_vis_shiny_app/data_marital.csv", row.names = FALSE)
```

```{r}
function_filter_1 <- function(df_var, df_data, cat_filter=NULL, df_clean_filtered) {

  df_sym <- sym(df_var)

  # Aggregate data for each decade and region
  df_simple <- map_dfr(
    names(decades),
    \(decade) {
      map_dfr(
        regions,
        \(reg) {
          df_clean_filtered %>%
            filter(
              year %in% decades[[decade]], region == reg) %>%
            group_by(sex, !!df_sym) %>%
            summarise(count = n(), .groups = "drop") %>%
            group_by(sex) %>%
            mutate(
              perc   = count / sum(count) * 100,
              region = reg,
              year   = decade
            )
        }
      )
    }
  )

  # Merge with external data and compute difference
  df_all <- df_data %>%
    left_join(df_simple, by = c("sex", df_var, "region", "year")) %>%
    mutate(
      count = replace_na(count, 0),
      perc  = replace_na(perc, 0)
    ) %>%
    arrange(year, region, sex, !!df_sym)

  df_all %>%
    select(-count) %>%
    pivot_wider(names_from = sex, values_from = perc) %>%
    mutate(percent = F - M)
}
```

# Checking data
```{r}
df_var <- "marital" #and educ
df_data <- df_marital

df_sym <- rlang::sym(df_var)  # convert string → symbol

df_simple <- purrr::map_dfr(
  names(decades),
  \(decade) {
    purrr::map_dfr(
      regions,
      \(reg) {
        df_clean %>%
          filter(year %in% decades[[decade]], region == reg) %>%
          group_by(sex, !!df_sym) %>%
          summarise(count = n(), .groups = "drop") %>%
          group_by(sex) %>%
          mutate(
            perc = count / sum(count) * 100,
            region = reg,
            year = decade
          ) 
      }
    )
  }
)

# Calculating percentage differences
df_all <- df_data %>%
  left_join(df_simple, by = c("sex", df_var, "region", "year")) %>%
  mutate(
    count = replace_na(count, 0),
    perc = replace_na(perc, 0)
  ) %>%
  arrange(year, region, sex, !!df_sym)

df_new <- df_all%>%
  filter(!count==0)

unique(df_clean$region)
```

#Inspecting distributions for percentage difference for all variables
```{r}
perc_happy <- function_filter_1(
      df_var  = "happiness",
      df_data = df_happy,
      df_clean_filtered = df_clean
    )

perc_educ <- function_filter(
      df_var  = "educ",
      df_data = df_educ
    )

perc_marital <- function_filter(
      df_var  = "marital",
      df_data = df_marital
    )


perc_happy <- perc_happy%>%
  #filter(percent !=0)
  filter(region== "Middle Atlantic" | region=="East North Central" | region=="New England" | region=="West North Central")

perc_educ <- perc_educ%>%
  #filter(percent !=0)
  filter(region== "Middle Atlantic" | region=="East North Central" | region=="New England" | region=="West North Central")

perc_marital <- perc_marital%>%
  #filter(percent !=0)
  filter(region== "Middle Atlantic" | region=="East North Central" | region=="New England" | region=="West North Central")


library(ggplot2)

happy_plot <- ggplot(perc_happy, aes(x = percent)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "steelblue") +
  scale_x_continuous(breaks = seq(floor(min(perc_happy$percent)),
                                  ceiling(max(perc_happy$percent)),
                                  by = 1)) +
  geom_density(alpha = 0.3) +
  labs(
    title = "Values for the variable Happiness",
    x = "Percentage difference between female and male scores",
    y = "Count"
  ) +
  theme_minimal()

educ_plot <- ggplot(perc_educ, aes(x = percent)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "steelblue") +
  scale_x_continuous(breaks = seq(floor(min(perc_educ$percent)),
                                  ceiling(max(perc_educ$percent)),
                                  by = 1)) +
  geom_density(alpha = 0.3) +
  labs(
    title = "Values for the variable Educational level",
    x = "Percentage difference between female and male scores",
    y = "Count"
  ) +
  theme_minimal()

marital_plot <- ggplot(perc_marital, aes(x = percent)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "steelblue") +
  scale_x_continuous(breaks = seq(floor(min(perc_marital$percent)),
                                  ceiling(max(perc_marital$percent)),
                                  by = 1)) +
  geom_density(alpha = 0.3) +
  labs(
    title = "Values for the variable Marital",
    x = "Percentage difference between female and male scores",
    y = "Count"
  ) +
  theme_minimal()

list_all <- c(
  perc_happy$percent,
  perc_educ$percent,
  perc_marital$percent
)

list_all <- as.data.frame(list_all)

all_plot <- ggplot(list_all, aes(x = list_all)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "steelblue") +
  scale_x_continuous(breaks = seq(floor(min(list_all)),
                                  ceiling(max(list_all)),
                                  by = 1)) +
  geom_density(alpha = 0.3) +
  labs(
    title = "Values for all variables combined",
    x = "Percentage difference between female and male scores",
    y = "Count"
  ) +
  theme_minimal()

happy_plot
educ_plot
marital_plot
all_plot
```

```{r}
??ggplotly
```

```{r}
demo("crosstalk-highlight-ggplotly", package = "plotly")
```
```{r}
help(highlight_key)
```


```{r}
d <- highlight_key(txhousing, ~city)
p <- ggplot(d, aes(date, median, group = city)) + geom_line()
gg <- ggplotly(p, tooltip = "city") 
highlight(gg, dynamic = TRUE)

# supply custom colors to the brush 
cols <- toRGB(RColorBrewer::brewer.pal(3, "Dark2"), 0.5)
highlight(gg, on = "plotly_hover", color = cols, dynamic = TRUE)

# Use attrs_selected() for complete control over the selection appearance
# note any relevant colors you specify here should override the color argument
s <- attrs_selected(
  showlegend = TRUE,
  mode = "lines+markers",
  marker = list(symbol = "x")
)

highlight(layout(gg, showlegend = TRUE), selected = s)
```


```{r}
# Example DeepSeek
library(plotly)
library(shiny)

ui <- fluidPage(
  fluidRow(
    column(4, plotlyOutput("hist_mpg")),
    column(4, plotlyOutput("hist_wt")),
    column(4, plotlyOutput("hist_cyl"))
  ),
  plotlyOutput("scatter_plot")
)

server <- function(input, output) {
  
  # Reactive for filtered data
  filtered_data <- reactiveVal(mtcars)
  
  # Observe clicks on any histogram
  observeEvent(event_data("plotly_click"), {
    click <- event_data("plotly_click")
    if (!is.null(click)) {
      # Simple filtering logic - adjust based on your needs
      current_data <- mtcars
      if (click$curveNumber == 0) { # MPG histogram
        filtered_data(current_data[current_data$mpg >= click$x, ])
      } else if (click$curveNumber == 1) { # WT histogram
        filtered_data(current_data[current_data$wt >= click$x, ])
      }
    }
  })
  
  output$hist_mpg <- renderPlotly({
    plot_ly(filtered_data(), x = ~mpg, type = "histogram") %>%
      layout(title = "MPG")
  })
  
  output$hist_wt <- renderPlotly({
    plot_ly(filtered_data(), x = ~wt, type = "histogram") %>%
      layout(title = "Weight")
  })
  
  output$hist_cyl <- renderPlotly({
    plot_ly(filtered_data(), x = ~cyl, type = "histogram") %>%
      layout(title = "Cylinders")
  })
  
  output$scatter_plot <- renderPlotly({
    plot_ly(filtered_data(), x = ~mpg, y = ~wt, type = 'scatter', mode = 'markers')
  })
}

shinyApp(ui, server)


```


```{r}
# Reactive value to store filtered data
filtered_data <- reactiveVal(df_clean)

# When year is clicked, update the reactive value
observeEvent(event_data("plotly_click", source = "year"), {
  click_data <- event_data("plotly_click", source = "year")
  
  if (!is.null(click_data)) {
    clicked_year_range <- click_data$x
    
    # Filter and store
    filtered_df <- df_clean %>% 
      filter(year_ranges == clicked_year_range)
    
    filtered_data(filtered_df)
  }
})

# Reset button
observeEvent(input$reset_filter, {
  filtered_data(df_clean)
})

# Plots use the reactive filtered data
output$bar_year <- renderPlotly({
  # Year plot always shows ALL data (not filtered by itself)
  p <- histogram_continuous(x = "year_ranges", title = "Year ranges", 
                           df = df_clean)
  ggplotly(p, source = 'year')
})

output$bar_genre <- renderPlotly({
  # Other plots use filtered data
  p <- histogram_continuous(x = "genre", title = "Genre Distribution", 
                           df = filtered_data())
  ggplotly(p, source = 'genre')
})

output$scatter_rating <- renderPlotly({
  p <- ggplot(filtered_data(), aes(x = rating, y = views)) +
    geom_point()
  ggplotly(p, source = 'rating')
})
```


# Example github
https://github.com/GtheSheep/Plotly-Shiny-Examples/blob/master/server.R

https://plotly-r.com/linking-views-with-shiny

# Part 3
## Run the app
```{r}
library(shiny)
# runApp("data_vis_shiny_app") #Shows only the app and no code

#runApp("data_vis_shiny_app", display.mode = "showcase") #Shows the code as well and highlights chinks with interaction
```